Two subtle problems with over-representation analysis

### 2024-06-10 manuscript/manuscript.Rmd
ms

RNA-seq counts for SRA run accession ERR2539161 were obtained from DEE2 [@Mo2018-zb].
This dataset was selected due to its high number of assigned reads (367M).
A library of 200 random gene sets were generated, each containing 30 genes.
To generate differential expression profiles, 5% of gene sets were selected to be
differentially expressed, with equal numbers of up- and down-regulated pathways.
Pseudosamples were generated by first downsampling the counts using the `thincounts`
function of the `edgeR` package v4.2.0 [@Robinson2010-eb] to 20M reads, followed by the
addition of some extra variation drawn from a normal distribution with mean of 1 and
standard deviation varying between 0.1 and 1.0, this value selected randomly for each gene.
Three "control" and five "case" pseudosamples were generated.
For the genes selected to be differentially expressed, expression values were multiplied by a
fold change in the case samples of 2.0 for upregulated genes and 0.5 for downregulated genes.
Genes that were selected to be both up and downregulated were left unchanged.
With these expression profiles, DESeq2 was used for differential analysis, and the results
were subjected to `enricher` which suffers from both problems, `fora`, an ORA function of
the `fgsea` bioconductor package v1.30.0 which does not suffer these problems, and `fgsea` a
fast implementation of the preranked GSEA algorithm [@Korotkevich2016-gd].
For the ORA functions above, significantly differentially expressed genes with FDR 0.05 were
selected for the foreground gene set.
If fewer than 200 genes are detected as significant, then 200 genes with the smallest p-values
were used for ORA, similar to a previous report [@Tarca2013-xh].
For `fgsea`, the DESeq2 test statistic was used for scoring differential expression.
After enrichment analysis with these approaches, the significant sets up and down regulated
sets were selected and compared to the ground truth in order to calculate precision, recall
and F1 score.
At each value of added variance, 1000 replications were conducted.
No set seed was used, but results were similar when run on different computer systems.

### Results

@@

@@

@@

>>>>

RNA-seq counts for SRA run accession ERR2539161 were obtained from DEE2 [@Mo2018-zb].
This dataset was selected due to its high number of assigned reads (367M).
A library of 200 random gene sets were generated, each containing 30 genes.
To generate differential expression profiles, 5% of gene sets were selected to be
differentially expressed, with equal numbers of up- and down-regulated pathways.
Pseudosamples were generated by first downsampling the counts using the `thincounts`
function of the `edgeR` package v4.2.0 [@Robinson2010-eb] to 20M reads, followed by the
addition of some extra variation drawn from a normal distribution with mean of 1 and
standard deviation varying between 0.1 and 1.0, this value selected randomly for each gene.
Three "control" and five "case" pseudosamples were generated.
For the genes selected to be differentially expressed, expression values were multiplied by a
fold change in the case samples of 2.0 for upregulated genes and 0.5 for downregulated genes.
Genes that were selected to be both up and downregulated were left unchanged.
With these expression profiles, DESeq2 was used for differential analysis, and the results
were subjected to `enricher` which suffers from both problems, `fora`, an ORA function of
the `fgsea` bioconductor package v1.30.0 which does not suffer these problems, and `fgsea` a
fast implementation of the preranked GSEA algorithm [@Korotkevich2016-gd].
For the ORA functions above, significantly differentially expressed genes with FDR 0.05 were
selected for the foreground gene set.
If fewer than 200 genes were detected as significant, then 200 genes with the smallest p-values
were used for ORA, similar to a previous report [@Tarca2013-xh].
For `fgsea`, the DESeq2 test statistic was used for scoring differential expression.
After enrichment analysis with these approaches, the significant up and down regulated
sets were selected and compared to the ground truth in order to calculate precision, recall
and F1 score.
At each value of added variance, 500 replications were conducted.
No set seed was used, but results were similar when run on different computer systems.

## Results

@@

Simulated differential expression profiles with apriori changes to predetermined pathway genes were
analysed in parallel by clusterProfilers enricher function, which suffers from the two aforementioned
problems and fora which does not.
In parallel, fgsea was used to compare the performance of these two ORA methods against a prototypical
FCS method.
ClusterProfiler scored very high precision when added variance was less than 0.4, but showed a
progressive worsening in precision as more noise was added to the dataset, dropping to 0.68 with added
variance of 1.0.
In comparison, the precision of fora was relatively stable around 0.97 until added variance reached 0.9.
The recall of ORA methods dropped suddenly when added variance was 0.5 or larger.
fgsea showed stable precision of 0.94 across all levels of added variance, and superior recall overall.

@@

Here we have described the effect of two subtle problems with common implementations of ORA.
This work was inspired by a doctoral student new to bioinformatics who was puzzled by confusing results
obtained from clusterProfiler when using a custom set of pathways.
Our hunch that clusterProfiler was behaving in an undesired way was confirmed by several posts to
online bioinformatics forums dating to 2020 where confused users were wondering why the number of
background genes changed so much.
ClusterProfiler maintainers recently provided an option to prevent the loss of unannotated genes from the
background `options(enrichment_force_universe = TRUE)`, however this feature isn't yet described in the
official documentation.

The second problem is potentially more serious in that it could lead to lots of false positives,
as clusterProfiler underestimates the actual number of tests that were conducted.
As shown in Figure 2B, this affects results differently depending on the gene set library used and also
the number of genes selected in the foreground list.

ClusterProfiler is not the only tool to suffer from these problems, they are widespread (**Table 1**).

| Tool | Has problems | Citation |
| --- | --- | --- |
| ClusterProfiler | Background and FDR | [@Yu2012-jj;@Wu2021-wy] |
| DAVID | Background and FDR | [@Dennis2003-ar;@Huang2009-gi;@Sherman2022-jn] |
| fora (fgsea) | Neither | [@Korotkevich2016-gd] |

: Table 1. Classification of several popular ORA tools for Background and FDR problems.

@@

The reason why such problems proliferated for so long is that using a custom background for ORA seems
to be the exception, rather than the rule.
Several tools do not even allow for the use of a custom background, suggesting that some tool developers
were unaware of the seriousness of this error.
This is one reason why ORA tool documentation also seems to lack any explanation of the importance of
using a custom background set.
